Nested Stochastic Block Models Applied to the Analysis of Single Cell Data

### 2021-07-20 manuscript.tex
final changes

To show this we took advantage of public spatial RNA dataset of a coronal section of murine brain tissue profiled with 10X Visium H\&E technology \cite{Gracia_Villacampa_2020}, as provided by the recently introduced package \texttt{SquidPy} \cite{Palla_Theis_2021}. We chose to stick to the given tissue annotation by the package authors. At default resolution, Leiden clustering resolves the tissue structure, as does the first level of the nSBM hierarchy (Fig. \ref{Figure_Visium}). When resolution is decreased (\emph{e.g.} $\gamma = 0.5$), the dentate gyrus is incorrectly merged to the hippocampus, whereas nSBM correctly identifies the pyramidal layer.

\begin{figure}[H]
\centering
\includegraphics[keepaspectratio,width=0.6\textwidth,height=\textheight]{Figure_Visium_Spatial.pdf}
\vspace*{-80pt}
\caption[]{Analysis of spatial transcriptomics of a coronal section of mouse brain by Visium H\&E. In the first panel, original tissue annotation is given. Tissues are well defined at default resolution for the standard approach. When resolution is decreased to $\gamma = 0.5$, cells from the dentate gyrus are merged to the hippocampus. When nSBM is used, the structure of the pyramidal is maintained at different levels of the hierarchy}\label{Figure_Visium}
\end{figure}

In another context, we tested the effect on the interpretation of the hierarchy varying the resolution parameter. We analysed data for hematopoietic differentiation \cite{paul_2015}, previously used to benchmark the consistency of cell grouping with differentiation trajectories by graph abstraction \cite{wolf_2019} (Fig. \ref{Figure_Hemato_Supp}A). Data show three major branchings (Erythroids, Neutrophils and Monocytes) stemming from the progenitor cells, mostly recapitulated by level 2 of the hierarchy computed by \texttt{schist} (Fig. \ref{Figure_Hemato_Lineage}). Not only the hierarchic model recapitulates the branching trajectories, also the cell groups appear to be consistent with the estimated pseudotime. Conversely, the Leiden method at default resolution identified 24 groups. By lowering the $\gamma$ parameter we observed cell groups that merge and split at different resolutions disrupting the hierarchy (Fig. \ref{Figure_Paul15_Leiden_r}).

@@

\begin{figure}[H]
\centering
\includegraphics[keepaspectratio,width=1\textwidth,height=\textheight]{Figure_Label_Transfer.pdf}
\vspace*{-40pt}
\caption[]{Label transfer using SBM. The first line reports UMAP embeddings for datasets profiled with Chromium 10X V3, Quartz-seq2 and  MARS-seq, each annotated by known cell types. Quartz-seq2 and MARS-seq were reannotated using \emph{k}NN method, \texttt{sc.tl.ingest() } or \texttt{schist}. The accuracy of each label transfer task is reported above the corresponding UMAP.} \label{Figure_Label_Transfer}
\end{figure}

@@

An alternative strategy is to evaluate the behaviour of modularity at different hierarchy levels. While nSBM does not optimise the graph modularity $Q$, we observed that this tends to be maximal for the level better describing known cell populations, so the optimal level $i^Q$ is

@@

As expected, the larger the network, the higher the optimal level. For relatively small datasets (\emph{i.e.} less than 10k cells), the first level of the hierarchy contains a number of groups in line with how many observable populations are. 
\textcolor{red}{Notwithstanding, cell groups identified at leach level may have a biological interpretation. In particular, groups at deepest levels (0 or 1) may be relevant when studying rare populations. For example, in the hematopoiesis dataset shown in Fig. \ref{Figure_Hemato_Supp}A, groups 11 (DC) and 19 (Lymph) cannot be distinguished from nSBM level 1 and up; a closer investigation to level 0, however, revealed that these cells are clearly separated. To better understand the role of deepest levels, we performed an additional analysis of a single cell dataset of mouse crypt cells \cite{Ayyaz_Kumar_2019}, which was also covered in a recent paper proposing GapClust as an optimal approach to identify rare cell populations \cite{Fa_Wei_Zhou_Johnston_2021}. We sought to identify the four rare populations identified by GapClust. We could identify all but the erythrocyte group (R3) at level 1 of the hierarchy (Fig. \ref{Figure_Rare_Main} and \ref{Figure_Rare_Supp}), suggesting that studying levels with appropriate community size is a valid method to spot rare populations. Of note, modularity optimisation could not identify Tuft cells in appropriate way (Fig. \ref{Figure_Rare_Supp}C), not even at high resolution, hence prompting the development of specific approaches such as GapClust.}

\begin{figure}[H]
\centering
\includegraphics[keepaspectratio,width=0.6\textwidth,height=\textheight]{Figure_Rare_Cells_Main.pdf}
%\vspace*{-40pt}
\caption[]{Identification of rare populations. FOOBAR} \label{Figure_Rare_Main}
\end{figure}

\textcolor{red}{As the size and number of communities is strictly dependent on the \emph{k}NN graph generation, we investigated how different parameters (\emph{i.e.} number of principal components and number of neighbors) affect the partition structure (Fig. \ref{Figure_Comm_Count} and \ref{Figure_Comm_Size}). We found, as a general pattern, that increasing the number of neighbors results in more granular structure at level 0, whereas increasing the number of PCs works in opposite direction. Looking at the agreement among different partitions, we found that it is consistent among different solutions, both at level 0 and level 1, with the exception of models for graph derived with low number of PC (Fig. \ref{Figure_Clustermap_L0}).}
%\subsection*{PPBM?}

@@

Minimisation of the nSBM is a process that requires a large amount of computational resources. While the underlying \texttt{graph-tool} library is efficient in exploring the solution space using a multiflip MCMC sampling strategy, the number of required iterations before convergence is considerable and the running time scales linearly with the number of edges. Moreover, to collect a consensus partition, we minimise multiple models (default: 100) that need to be averaged. To give a reference, we report runtimes for some example datasets in Table \ref{Table_runtime} on a commodity hardware (Intel i7@2.8 GHz, 32 GB RAM). Compared to Leiden approach, nSBM requires at best $\sim6\times$ times more, and $\sim30\times$  at worst. A reasonably fast alternative to the nSBM is the Planted Partition Block Model (PPBM), for which we also report runtimes. The PPBM \cite{Zhang_Peixoto_2020} is able to find statistically significant assortative modules and eliminates the resolution parameter; differently from nSBM, PPBM is not hierarchic.

>>>>

To show this we took advantage of public spatial RNA dataset of a coronal section of murine brain tissue profiled with 10X Visium H\&E technology \cite{Gracia_Villacampa_2020}, as provided by the recently introduced package \texttt{SquidPy} \cite{Palla_Theis_2021}. We chose to stick to the given tissue annotation by the package authors. At default resolution, Leiden clustering resolves the tissue structure, as does the first level of the nSBM hierarchy (Fig. \ref{Figure_Visium}). When resolution is decreased (\emph{e.g.} $\gamma = 0.5$), the dentate gyrus is incorrectly merged to the hippocampus, whereas \texttt{schist} correctly identifies the pyramidal layer.

\begin{figure}[H]
\centering
\includegraphics[keepaspectratio,width=0.6\textwidth,height=\textheight]{Figure_Visium_Spatial.pdf}
\vspace*{-80pt}
\caption[]{Analysis of spatial transcriptomics of a coronal section of mouse brain by Visium H\&E. In the first panel, original tissue annotation is given. Tissues are well defined at default resolution for the standard approach. When resolution is decreased to $\gamma = 0.5$, cells are aggregated breaking the histological types, \emph{e.g.} cells from the dentate gyrus are merged to the hippocampus. When a nSBM is applied, the structure of the pyramidal layer is maintained at different levels of the hierarchy}\label{Figure_Visium}
\end{figure}

In another context, we tested the effect on the interpretation of the hierarchy varying the resolution parameter. We analysed data for hematopoietic differentiation \cite{paul_2015}, previously used to benchmark the consistency of cell grouping with differentiation trajectories by graph abstraction \cite{wolf_2019} (Fig. \ref{Figure_Hemato_Supp}A). Data show three major branchings (Erythroids, Neutrophils and Monocytes) stemming from the progenitor cells, mostly recapitulated by level 2 of the hierarchy computed by \texttt{schist} (Fig. \ref{Figure_Hemato_Lineage}). Not only the hierarchic model recapitulates the branching trajectories, also the cell groups appear to be consistent with the estimated pseudotime (Fig. \ref{Figure_Hemato_Supp}B). Conversely, the Leiden method at default resolution identified 24 groups. By lowering the $\gamma$ parameter we observed cell groups that merge and split at different resolutions disrupting the hierarchy (Fig. \ref{Figure_Paul15_Leiden_r}).

@@

\begin{figure}[H]
\centering
\includegraphics[keepaspectratio,width=1\textwidth,height=\textheight]{Figure_Label_Transfer.pdf}
\vspace*{-40pt}
\caption[]{Label transfer using SBM. The first line reports UMAP embeddings for datasets profiled with Chromium 10X V3, Quartz-seq2 and  MARS-seq, each annotated by known cell types. Quartz-seq2 and MARS-seq were reannotated using \emph{k}NN method, \texttt{sc.tl.ingest()} or \texttt{schist}. The accuracy of each label transfer task is reported above the corresponding UMAP.} \label{Figure_Label_Transfer}
\end{figure}

@@

An alternative strategy is to evaluate the behaviour of modularity at different hierarchy levels. While \texttt{schist} does not optimise the graph modularity $Q$, we observed that this tends to be maximal for the level better describing known cell populations, so the optimal level $i^Q$ is

@@

As expected, the larger the network, the higher the optimal level. For relatively small datasets (\emph{i.e.} less than 10k cells), the first level of the hierarchy contains a number of groups in line with how many observable populations are. 
\textcolor{red}{Notwithstanding, cell groups identified at leach level may have a biological interpretation. In particular, groups at deepest levels (0 or 1) may be relevant when studying rare populations. For example, in the hematopoiesis dataset shown in Fig. \ref{Figure_Hemato_Supp}A, groups 11 (DC) and 19 (Lymph) cannot be distinguished from nSBM level 1 and up; a closer investigation to level 0, however, revealed that these cells are clearly separated (Fig. \ref{Figure_Hemato_Supp}D). To better understand the role of deepest levels, we performed an additional analysis of a single cell dataset of mouse crypt cells \cite{Ayyaz_Kumar_2019}, which was also covered in a recent paper proposing GapClust as an optimal approach to identify rare cell populations \cite{Fa_Wei_Zhou_Johnston_2021}. We sought to identify the four rare populations identified by GapClust. We could distinguish all but the erythrocyte group (R3) at level 1 of the hierarchy (Fig. \ref{Figure_Rare_Main} and \ref{Figure_Rare_Supp}), suggesting that exploring nSBM levels with appropriate community size is a valid method to spot rare populations. Of note, modularity optimisation could not pinpoint Tuft cells in appropriate way (Fig. \ref{Figure_Rare_Supp}C), not even at high resolution, hence prompting the development of specific approaches such as GapClust.}

\begin{figure}[H]
\centering
\includegraphics[keepaspectratio,width=0.6\textwidth,height=\textheight]{Figure_Rare_Cells_Main.pdf}
%\vspace*{-40pt}
\caption[]{Identification of rare populations. UMAP embeddings of mouse crypt cells highlighting the groups found at level 1 of nSBM hierarchy corresponding to rare populations identified by GapClust \cite{Fa_Wei_Zhou_Johnston_2021}. Each rare population is colored by a signature calculated from published gene lists (see methods).} \label{Figure_Rare_Main}
\end{figure}

\textcolor{red}{As the size and number of communities is strictly dependent on the \emph{k}NN graph generation, we investigated how different parameters (\emph{i.e.} number of principal components and number of neighbors) affect the partition structure (Fig. \ref{Figure_Comm_Count} and \ref{Figure_Comm_Size}). We found, as a general pattern, that increasing the number of neighbors results in more granular structure at level 0, with different solutions being consistent (Fig. \ref{Figure_Clustermap_L0}), suggesting that higher number of neighbors provides richer description of the dataset. The number of PCs used to evaluate cell-to-cell distance influences the variability of community sizes; the consistency among different solutions is high when a sufficient number of PCs is chosen, data suggest that for large datasets more PCs should be included to include adequate fraction of overall variability.}
%\subsection*{PPBM?}

@@

Minimisation of a nSBM is a process that requires a large amount of computational resources. While the underlying \texttt{graph-tool} library is efficient in exploring the solution space using a multiflip MCMC sampling strategy, the number of required iterations before convergence is considerable and the running time scales linearly with the number of edges. Moreover, to collect a consensus partition, we minimise multiple models (default: 100) that need to be averaged. To give a reference, we report runtimes for some example datasets in Table \ref{Table_runtime} on a commodity hardware (Intel i7@2.8 GHz, 32 GB RAM). Compared to Leiden approach, nSBM requires at best $\sim6\times$ times more, and $\sim30\times$  at worst. A reasonably fast alternative to the nSBM is the Planted Partition Block Model (PPBM), for which we also report runtimes. The PPBM \cite{Zhang_Peixoto_2020} is able to find statistically significant assortative modules and eliminates the resolution parameter; differently from nSBM, PPBM is not hierarchic.

### 2021-07-20 supplementary.tex
final changes

\section*{Supplementary tables}
\begin{table}[h!]
\centering
 \begin{tabular}{||c|c c|c c||}
%  \begin{tabular}{|| m{7em} | m{5em} | m{7em} | m{7em} | m{5em} ||}
 \hline
\multicolumn{1}{||c|}{\textbf{Noise level ($\sigma$)}} & \multicolumn{2}{c|}{\textbf{Leiden}} & \multicolumn{2}{c||}{\textbf{nSBM level 1}} \\

 & $ARI$ & $N$ & $ARI$ & $N$  \\ 
 \hline\hline
-- & 0.9228 & 8 & 0.8944 & 6 \\
\hline 
0.5 & 0.6249 & 10 & 0.7509  & 10 \\ 
 \hline
 1.0  & 0.5966 & 7 & 0.7269  & 6 \\
 \hline
 1.5  & 0.2833 & 9 &0.3029 & 3 \\
 \hline
 2.0  & 0.1405 & 12 & 0.1961 & 2 \\
 \hline
\end{tabular}
\caption{\textcolor{red}{Effect of random noise }}
\label{Table_Noise}
\end{table}

@@

\begin{figure}[htbp]
\centering
\includegraphics[keepaspectratio,width=\textwidth,height=0.375\textheight]{Random_Leiden_by_gamma.png}
\caption[]{\textcolor{red}{foobarbaz}}
\label{Modularity_Random_Res}
\end{figure}
\clearpage

@@

\begin{figure}[htbp]
\centering
\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Figure_Hemato_Supp.pdf}
\caption[]{Analysis of hematopoietic differentiation. (A) Low dimensional embedding of single cells coloured by original cell type and pseudotime. (B) Cells are coloured according to the nSBM grouping at level \textcolor{red}{1} of the hierarchy, next to a radial tree representation of the same model.}\label{Figure_Hemato_Supp}
\end{figure}
\clearpage

@@

\clearpage
\begin{figure}[H]
\centering
\includegraphics[keepaspectratio,width=1\textwidth,height=\textheight]{Figure_Rare_Cells_Supp.pdf}
%\vspace*{-40pt}
\caption[]{Identification of rare populations. FOOBAR} \label{Figure_Rare_Supp}
\end{figure}

@@

%\vspace*{-40pt}
\caption[]{Consistency at Level 0} \label{Figure_Clustermap_L0}
\end{figure}

>>>>

\section*{Supplementary tables}
\begin{table}[h!]
\centering
 \begin{tabular}{||c|c c|c c||}
%  \begin{tabular}{|| m{7em} | m{5em} | m{7em} | m{7em} | m{5em} ||}
 \hline
\multicolumn{1}{||c|}{\textbf{Noise level ($k$)}} & \multicolumn{2}{c|}{\textbf{Leiden}} & \multicolumn{2}{c||}{\textbf{nSBM level 1}} \\

 & $ARI$ & $N$ & $ARI$ & $N$  \\ 
 \hline\hline
-- & 0.9228 & 8 & 0.8944 & 6 \\
\hline 
0.5 & 0.6249 & 10 & 0.7509  & 10 \\ 
 \hline
 1.0  & 0.5966 & 7 & 0.7269  & 6 \\
 \hline
 1.5  & 0.2833 & 9 &0.3029 & 3 \\
 \hline
 2.0  & 0.1405 & 12 & 0.1961 & 2 \\
 \hline
\end{tabular}
\caption{Effect of random noise. The table shows performance of Leiden approach and nested Stochastic Block Model in recognising annotated population in PBMC 3k data when increasing level of noise are added to log-normalized counts. The $k$ parameter multiplies the $\sigma$, estimated from data, to generate random counts. $ARI$: Adjusted Rand Index, $N$: number of communities.}
\label{Table_Noise}
\end{table}

@@

\begin{figure}[htbp]
\centering
\includegraphics[keepaspectratio,width=\textwidth,height=0.375\textheight]{Random_Leiden_by_gamma.png}
\caption[]{Decreasing the resolution parameter $\gamma$ results in lower number of communities also in the random graph generated from PBMC 3k data. At $\gamma < 0.5$, Leiden approach correctly identifies one single community.}
\label{Modularity_Random_Res}
\end{figure}
\clearpage

@@

\begin{figure}[htbp]
\centering
\includegraphics[keepaspectratio,width=\textwidth,height=0.75\textheight]{Figure_Hemato_Supp.pdf}
\caption[]{Low dimensional embedding of single cells from hematopoietic differentiation dataset. Cells are colored by original cell type (A), pseudotime (B), nSBM grouping at level \textcolor{red}{1} of the hierarchy (C) and level 0 (D). Since level 0 identifies 114 groups, only group labels have been represented.}\label{Figure_Hemato_Supp}
\end{figure}
\clearpage

@@

\clearpage
\begin{figure}[H]
\centering
\includegraphics[keepaspectratio,width=1\textwidth,height=\textheight]{Figure_Rare_Cells_Supp.pdf}
%\vspace*{-40pt}
\caption[]{Identification of rare populations. (A) UMAP embedding of mouse crypt cells coloured by cell groups resulting at level 1 of the nSBM hierarchy. (B) Cells coloured by the signature of R3 group identified by GapClust. (C) UMAP embedding of mouse crypt cells coloured after optimisation of modularity at default parameters (left) or higher resolution (right).} \label{Figure_Rare_Supp}
\end{figure}

\clearpage
\begin{figure}[H]
\centering
\includegraphics[keepaspectratio,width=.95\textwidth,height=\textheight]{Community_Count_L0.pdf}
%\vspace*{-40pt}
\caption[]{Evaluation of the number of communities. Barplots show the number of communities found at level 0 of the nSBM hierarchy for various datasets used in the manuscript when different parameters are chosen to build the \emph{k}NN graph. Increasing the number of neighbors results in more granular partitioning (\emph{i.e.} more communities), whereas increasing the number of principal components used to evaluate cell-to-cell distance generally works in opposite way. $x$-axis: number of principal components; $y$-axis: number of communities; bar color: number of \emph{k}NN neighbors.} \label{Figure_Comm_Count}
\end{figure}

\clearpage
\begin{figure}[H]
\centering
\includegraphics[keepaspectratio,width=.95\textwidth,height=\textheight]{Community_Size_L0.pdf}
%\vspace*{-40pt}
\caption[]{Evaluation of the size of communities. Boxplots show the distribution of size of communities found at level 0 of the nSBM hierarchy for various datasets used in the manuscript when different parameters are chosen to build the \emph{k}NN graph. Increasing the number of neighbors results in more granular partitioning (\emph{i.e.} smaller communities), increasing the number of principal components results in more dispersed sizes. $x$-axis: number of principal components; $y$-axis: community size; bar color: number of \emph{k}NN neighbors.} \label{Figure_Comm_Size}
\end{figure}

@@

%\vspace*{-40pt}
\caption[]{Consistency at Level 0. Each heatmap shows the Adjusted Rand Index among partitions found when different parameters are used to build th \emph{k}NN graph. Row labels encode the number of PCs and the number of neighbors (P\_N), reflected in colors used to annotate columns.} \label{Figure_Clustermap_L0}
\end{figure}
